---
title: "Intro to GNU Make (part 1)"
subtitle: "Automating your workflow"
format: stat159slides-revealjs
chalkboard: true
---

# Example: Compound Interest

## Motivation

- You deposit $1000 in a savings account 
- The annual interest rate is 2%
- Assume interest rate won't change
- How much money will you have one year from now?

. . . 

The answer to this question is given by the [compound interest]{.bold-hilit} formula:

$$
\text{deposit} + \text{annual paid interest} = \text{amount in one year}
$$

. . .

$$
1000 + 1000 (0.02) = 1000 \times (1 + 0.02) = 1020
$$


## Motivation (cont'd)

How much money will you have 2 years from now?

. . .

$$
\text{amount year 1} + \text{annual paid interest} = \text{amount in year 2}
$$

. . .

$$
1020 + 1020 (0.02) = 1020 \times (1 + 0.02) = 1040.40
$$

\

. . .

How much money will you have in 3 years?

. . .

$$
\text{amount year 2} + \text{annual paid interest} = \text{amount in year 3}
$$

. . .

$$
1040.40 + 1040.40 (0.02) = 1040.40 \times (1 + 0.02) = 1061.208
$$


## Motivation (cont'd)

In one year you'll have:

$$
1000 \times (1 + 0.02) = 1020
$$

\

In two years you'll have:

$$
1000 \times (1 + 0.02) \times (1 + 0.02) = 1000 \times (1 + 0.02)^2 = 1040.4
$$

\

In three years you'll have:

$$
1000 \times (1 + 0.02) \times (1 + 0.02) \times (1 + 0.02) = 1000 \times (1 + 0.02)^3 = 1061.208
$$


## Future Value

In $n$ years you'll have:

$$
1000 \times (1 + 0.02)^n
$$

\

::: {.fragment}
### Future Value formula

$$
FV = P \times (1 + r)^n
$$

- $FV$ = Future Value
- $P$ = Principal (the initial deposit)
- $r$ = Annual rate of return (interest rate)
- $n$ = Number of years
:::


## Example A {auto-animate="true"}

```{r}
#| echo: true
# Future value (compound interest)
# amount in year 10
P = 1000
r = 0.05
n = 10
```


## Example A {auto-animate="true"}

```{r}
#| echo: true
#| output-location: fragment
# Future value (compound interest)
# amount in year 10
P = 1000
r = 0.05
n = 10

FV = P * (1 + r)^n
FV
```



## Example B {auto-animate="true"}

```{r}
#| echo: true
# Future value (compound interest)
# amount in every year 1-10
P = 1000
r = 0.05
n = 1:10
```


## Example B {auto-animate="true"}

```{r}
#| echo: true
#| output-location: fragment
# Future value (compound interest)
# amount in every year 1-10
P = 1000
r = 0.05
n = 1:10

FV = P * (1 + r)^n
FV
```



## Example B {auto-animate="true"}

:::: {.columns}
::: {.column}
```{r}
#| echo: true
#| output-location: fragment
# Future value
# amount in every year 1-10
P = 1000
r = 0.05
n = 1:10

FV = P * (1 + r)^n

future_values = data.frame(
  year = n, 
  amount = FV)
```
:::

::: {.column .fragment}
```{r}
#| echo: true
future_values
```
:::
::::

## Example B {auto-animate="true"}

```{r}
#| echo: true
#| output-location: fragment
library(ggplot2)

ggplot(data = future_values, aes(x = year, y = amount)) +
  geom_col() +
  theme_bw()
```


# Sample Project

## Sample project `future_value`

- Suppose we have a directory `future_value/`
- This directory contains a quarto file `doc.qmd`
- The qmd file has code to compute future value (based on our working example)

. . .

Assuming our working directory is `future_value`, we can generate an output
file `doc.html` by running the following command on the terminal:

```{.bash code-line-numbers="false"}
# create output file: doc.html
quarto render doc.qmd
```


##

File structure (at the beginning)

```{.markdown code-line-numbers="false"}
future_value/
  doc.qmd
```

. . .

\

Run command in terminal

```{.bash code-line-numbers="false"}
# create output file: doc.html
quarto render doc.qmd
```

. . .

\

File structure (at the end)

```{.markdown code-line-numbers="false"}
future_value/
  doc.qmd
  doc.html
```


# GNU Make

##

File structure (at the beginning)

```{.markdown code-line-numbers="false"}
future_value/
  doc.qmd
  Makefile
```

. . .

\

Run command in terminal

```{.bash code-line-numbers="false"}
# let Make do its magic
make
```

. . .

\

File structure (at the end)

```{.markdown code-line-numbers="false"}
future_value/
  doc.qmd
  doc.html
  Makefile
```

## About GNU Make

- GNU Make, commonly known as [make]{.bold-hilit}, is a Unix tool that is part of 
the [GNU Project](https://www.gnu.org/software/).

- To work with Make you write a text file with the name `Makefile`, 
typically referred to as the _makefile_.

- Notice that `Makefile` has not file extensions!


## What's in a Makefile?

A _makefile_ is formed by [rules]{.bold-hilit} that have the following structure:

. . .

```{.markdown filename="Makefile" code-line-numbers="false"}
# rule
target: dependencies
  commands
```

\

- A rule tells Make how to execute a series of commands in order to build a 
target file from source files (i.e. the dependencies). 

- ⚠️ IMPORTANT: the recipe line(s) must be [indented with a TAB]{.bold-hilit} and 
not with any spaces.


## Basic rule syntax

. . .

Under its most basic form, a rule usually has this structure:

```{.markdown filename="Makefile" code-line-numbers="false"}
# rule to generate target
target: dependency
  command
```

\

- `target` is the output file
- `dependency` is the input file
- `command` is the program that creates the `target`
- the `command` line is [indented with a TAB]{.bold-hilit}


## Let's go back to our future value example {auto-animate="true"}

. . .

```{.markdown filename="Makefile" code-line-numbers="false"}
# recipe to generate output file doc.html
```

## Let's go back to our future value example {auto-animate="true"}

```{.markdown filename="Makefile" code-line-numbers="false"}
# recipe to generate output file doc.html
doc.html: doc.qmd
```


## Let's go back to our future value example {auto-animate="true"}

```{.markdown filename="Makefile" code-line-numbers="false"}
# recipe to generate output file doc.html
doc.html: doc.qmd
  quarto render doc.qmd
```

\

- `doc.html`: [target]{.hilit} or output file
- `doc.qmd`: [dependency]{.mdlit} or input file
- `quarto render doc.qmd`: [command]{.bold-hilit} to create [target]{.hilit}


## Generate PDF output

What if we wanted to generate an output document in PDF format?

. . . 

```{.markdown filename="Makefile" code-line-numbers="false"}
# recipe to generate output file doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf
```

. . .

\

::: {.nonincremental}
- `doc.pdf`: [target]{.hilit} or output file
- `doc.qmd`: [dependency]{.mdlit} or input file
- `quarto render doc.qmd --to pdf`: [command]{.bold-hilit} to create [target]{.hilit}
:::


# Various Rules

## More than one rule {auto-animate="true"}

Say we want to create 2 outputs: HTML and PDF docs

. . .

```{.markdown filename="Makefile" code-line-numbers="false"}
# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd
```

## More than one rule {auto-animate="true"}

Say we want to create 2 outputs: HTML and PDF docs

```{.markdown filename="Makefile" code-line-numbers="false"}
# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf
```

## More than one rule 

Say we want to create 2 outputs: HTML and PDF docs

```{.markdown filename="Makefile" code-line-numbers="false"}
# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf
```

Running make will generate both outputs

```{.bash code-line-numbers="false"}
# run make to generate both outputs
make
```

## More than one rule 

Say we want to create 2 outputs: HTML and PDF docs

```{.markdown filename="Makefile" code-line-numbers="false"}
# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf
```

. . .

But we can generate single outputs, e.g. 

```{.bash code-line-numbers="false"}
make doc.html
```

\

. . .

```{.bash code-line-numbers="false"}
make doc.pdf
```


# Phony Targets

## Phony Targets

- By default, Makefile targets are [file targets]{.bold-hilit}

- This means that targets are used to build files from other files (i.e. the dependencies)

- This implies that Make assumes the target of a rule is a file.

- There are occasions, however, when a given target is not really a file, but 
simply the name of a recipe

- The most common example is the `all` target


## Phony target `all`

. . .

```{.markdown filename="Makefile" code-line-numbers="false"}
# all outputs
all: doc.html doc.pdf

# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf
```

- The first target `all` is not the name of a file; 

- It is simply a convenient “label” to indicate a series of dependencies. 

- This type of target is known in Make terminology as a [phony]{.bold-hilit} target.

- _Phony_ meaning "not a real file".


## Phony target `all` {auto-animate="true"}

The convention is to explicitly declare a phony target by making it a 
prerequisite of the special target `.PHONY` as follows:

. . .

```{.markdown filename="Makefile" code-line-numbers="false"}
# declaring phony target
.PHONY: all
```

## Phony target `all` {auto-animate="true"}

The convention is to explicitly declare a phony target by making it a 
prerequisite of the special target `.PHONY` as follows:

```{.markdown filename="Makefile" code-line-numbers="false"}
# declaring phony target
.PHONY: all

# all outputs
all: doc.html doc.pdf
```

## Phony target `all` {auto-animate="true"}

The convention is to explicitly declare a phony target by making it a 
prerequisite of the special target `.PHONY` as follows:

```{.markdown filename="Makefile" code-line-numbers="false"}
# declaring phony target
.PHONY: all

# all outputs
all: doc.html doc.pdf

# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf
```


## Declaring phony targets

Why do we declare phony targets?

. . .

\

Declaration of phony targets is important because of two main reasons:

1) to avoid a conflict with a file of the same name, and

2) to improve performance


## Phony target `clean` 

- Another common phony target is the so-called `clean` target

- This is used to indicate a recipe for "cleaning" a directory from secondary files 
typically generated after some compilation process.

- Also, the `clean` target is also commonly used to remove output files.

. . .

::: {.small}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean

all: <targets>

# rule 1
...

# rule 2
...

# remove output files
clean:
  rm -f <outputs>
```
:::


## Phony target `clean` {auto-animate="true"}

. . .

Here's what our `Makefile` could look like with phony targets `all` and `clean:`

::: {.small .fragment}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean

all: doc.html doc.pdf

# rule 1: doc.html
doc.html: doc.qmd
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd
  quarto render doc.qmd --to pdf

clean:
  rm -f doc.html doc.pdf
```
:::

. . .

::: {.small}
Note that `clean` has no dependencies. Therefore, the target `clean` will be always 
out-of-date. So whenever you run `make clean`, it will run independent from the 
state of the file system.
:::


# Various Dependencies

## More than one input

Suppose we have an R script file `functions.R` with the code of the 
future value formula.

. . . 

```{.markdown filename="functions.R" code-line-numbers="false"}
# Title: Future Value (compound interest)
# Inputs:
# - p: principal
# - r: annual rate of return
# - n: number of years
future_value <- function(p, r, n) {
  fv = p * (1 + r)^n
  return(fv)
}
```

. . .

And assume that `doc.qmd` "depends" on `functions.R`. In other words,
there's a code cell in the qmd file that loads the code in `functions.R`


## More than one input

The file structure now looks like this:

. . .

```{.markdown code-line-numbers="false"}
future_value/
  functions.R
  doc.qmd
  Makefile
```

\

. . .

Technically speaking:

- `doc.html` now depends on both `doc.qmd` and `functions.R`
- `doc.pdf` also depends on both `doc.qmd` and `functions.R`
- If either `functions.R` or `doc.qmd` change, then the output files 
should be updated


## More than one input {auto-animate="true"}

. . .

Here's what our `Makefile` could look like:

::: {.small .fragment}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean
```
:::


## More than one input {auto-animate="true"}

Here's what our `Makefile` could look like:

::: {.small}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean

all: doc.html doc.pdf
```
:::


## More than one input {auto-animate="true"}

Here's what our `Makefile` could look like:

::: {.small}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean

all: doc.html doc.pdf

# rule 1: doc.html
doc.html: doc.qmd functions.R
  quarto render doc.qmd
```
:::


## More than one input {auto-animate="true"}

Here's what our `Makefile` could look like:

::: {.small}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean

all: doc.html doc.pdf

# rule 1: doc.html
doc.html: doc.qmd functions.R
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd functions.R
  quarto render doc.qmd --to pdf
```
:::

## More than one input {auto-animate="true"}

Here's what our `Makefile` could look like:

::: {.small}
```{.markdown filename="Makefile" code-line-numbers="false"}
.PHONY: all clean

all: doc.html doc.pdf

# rule 1: doc.html
doc.html: doc.qmd functions.R
  quarto render doc.qmd

# rule 2: doc.pdf
doc.pdf: doc.qmd functions.R
  quarto render doc.qmd --to pdf

clean:
  rm -f doc.html doc.pdf
```
:::
