---
title: "10 Simple Rules for RR"
subtitle: "An Overview of Ten Simple Rules for Reproducible Research"
format: stat159slides-revealjs
chalkboard: true
---

# Ten Simple Rules for Reproducible Computational Research

## Reading

::: {.nonincremental}
- Title: **Ten Simple Rules for Reproducible Computational Research**
- Author(s): Sandve G.K., Nekrutenki A., Taylor J., Hovig E.
- Year: 2013
- Journal: PLOS Computational Biology 9(10): e1003285
- <https://doi.org/10.1371/journal.pcbi.1003285>
:::


## Why are we here?

- [Beyond Morality:]{.bold-hilit} Reproducibility isn't just a moral obligation to your field; 
a lack of tracking is a direct technical burden on *you*.

- [The Time-Saver:]{.bold-hilit} Meticulous tracking allows your future self to effortlessly 
apply past methods to new data or reuse code templates across separate projects.

- [The Bare-Minimum Threshold:]{.bold-hilit} At an absolute baseline, you must at least be able 
to reproduce your *own* results to defend your science against future questioning.

- [The Reality Check:]{.bold-hilit} Because any given analysis contains an exponential 
combination of parameters and software variables, memory *will* fail you. You must record details while underway.


<!--
## The Core Definition

An *analysis workflow* is the full sequence of:

- pre-processing, 
- script executions, 
- and post-processing required 
- to get from raw data to a final plot.
-->


## 

### Rules 1 & 2: Eliminating Human Drift

- [The Documentation Trap:]{.bold-hilit} Manual logs easily drift out of sync.

- [The Executable Solution:]{.bold-hilit} Write your entire workflow as a directly 
executable file—such as a shell script or a Makefile.

- [Ban Manual Clicks:]{.bold-hilit} Avoid manual copy-pasting, mouse clicks or graphical format 
tweaks between documents.

::: {.fragment}
### Rules 3 & 4: Freezing Your Software State

- [The Version Lock:]{.bold-hilit} Minor changes between software updates can alter data 
outputs or prevent code from running entirely.
:::


## Rule 5: Exposing the Pipeline's Inner Mechanics

\

::: {.columns}
::: {.column width="50%"}
### The Analytical Value

Quickly browsing through step-by-step intermediate output files can reveal 
silent logical discrepancies, catching bugs that are completely masked in 
final summary charts.
:::

::: {.column width="50%" .fragment}
### The Pipeline Value

If your full project execution is long or resource-heavy, archiving standardized 
intermediate states allows you to re-run localized parts of the process without 
starting over from scratch.
:::
:::

- [The Audit Baseline:]{.bold-hilit} Storing intermediate files allows external peers 
to critically examine your complete pipeline logic without needing all 
your local executables operational.


## Rules 6 & 7: Perfect Replications & Data Foundations

::: {.callout-warning}
### Rule 6: The Random Seed Mandate

Many analyses involve stochastic element randomness. To achieve an *exact* 
procedural reproduction rather than just an approximation, you must explicitly 
record and pass your initial random seed.
:::

- [Isolate Visual Tweaks:]{.bold-hilit} Store the data behind your charts.

- [The Scripted Plot:]{.bold-hilit} Preserve the exact code used to render a graphic.


<!--
## Rule 8: Generating Hierarchical Output

- **The Summary Illusion:** Final publication charts and tables represent heavily 
compressed, summarized distributions (like a single mean value along a curve).

- **The Inspection Obligation:** Fully validating your findings requires you to 
inspect and verify the granular, detailed sub-values underlying those summaries 
at least once.

- **The Hypertext Blueprint:** The authors advocate for structuring output as 
HTML hypertext. 

- **The Click Connection:** This creates an intuitive data ecosystem where 
readers can simply click a summarized data point to follow a link directly 
to its raw, underlying source matrix.


## Rule 9: Connecting Textual Claims to Result IDs

- **The Separation Flaw:** Throughout a project, interpretations live in separate 
text files, emails, or markdown notes, while data results live on a local 
workstation or cluster server.

- **The Matching Failure:** Tracing which exact version of an analysis supports 
a specific paragraph or conclusion becomes highly error-prone as the project's 
data pool scales.

- **The Early Bind:** Force yourself to embed explicit file paths, hash tags, 
or analysis result IDs directly into your notes the exact moment you draft 
an interpretation.

- **Literate Documents:** Leverage integrated processing engines (like Sweave, 
Quarto, or interactive workflow environments) to weld your live data results 
directly inside your descriptive text files.
-->


## Rule 10: Providing Public Access to Your Workflow

- [The Minimum Material:]{.bold-hilit} Submit your core data frames and raw 
source code scripts as mandatory supplementary online files upon publication.

- [The Review Advantage:]{.bold-hilit} Exposing your full workspace infrastructure 
increases reviewer trust, accelerates editorial turnaround times, and boosts your 
project's future citation metrics.

- [The Open Stance:]{.bold-hilit} Always remain prepared to openly answer peer 
requests for deep architectural or parameter pipeline metrics.


# Discussion Time

## Screens Off, Minds On

\

::: {.nonincremental}
- [Devices Away:]{.bold-hilit} Laptops & tablets closed, phones away.

- [No One Alone:]{.bold-hilit} Pair or team up (no solo seats).

- [Engage Fully:]{.bold-hilit} Be ready to speak, debate, and share ideas.
:::


## Scenario 1

Your teammate creates a breathtaking, beautifully styled 
bar chart for a group presentation. However, when the professor asks 
to see the 10,000 raw data points that generated the third bar, your 
teammate sweats and says, _"Uh, it's on my old laptop."_

. . .

::: {.poll .nonincremental}
- Can a highly styled chart have any actual academic value?

- How should losing access to raw files be treated?

- Should the student be accused of data fabrication?
:::

```{r}
#| echo: false
countdown::countdown(4, bottom = 0)
```

::: {.notes}
A pretty chart without raw data trails isn't science; it's just creative writing/coding.
There is not a lot of academic value in a breathtaking, highly styled chart 
if the baseline data supporting it cannot be produced on demand.

"The burden of proof." In some professional environments, an inability to instantly 
back up a specific visual claim with its underlying numbers looks less like a 
mistake and more like data fabrication.
:::


## Scenario 2

A student keeps a beautiful, 100-page handwritten leather notebook. 
She tracks every single mouse click, every copy-paste action, and every manual 
data-cleaning step she performs in Excel with flawless precision. 

. . .

::: {.poll .nonincremental}
- Is the student's work reproducible?

- Is there any value in (or place for) hand-written notes?
:::

```{r}
#| echo: false
countdown::countdown(4, bottom = 0)
```

::: {.notes}
A beautifully kept manual record still fails the core requirement of being an 
"executable description" that a computer can run and scale.

Manual copy-paste actions and Excel mouse clicks introduce silent, 
untraceable human errors that version control cannot catch.
We should prefer automated scripts that require zero human intervention to repeat.

There is immense value in handwritten notes and manual approaches, but only 
as a sandbox for human thought, not as the final pipeline for data execution.

The "Why" vs. the "How": Code is excellent at tracking how a calculation is done, 
but it is terrible at explaining why you chose to do it. 
A handwritten notebook is a cognitive sandbox. 
It is where a researcher maps out messy conceptual logic, sketches potential 
workflows, and processes failures before writing a single line of script.
:::


## Scenario 3

You spend three weeks testing a hypothesis on a dataset. 
Your code is perfect, your pipeline is automated, but the results prove 
your hypothesis is completely, totally wrong. Your script yields a massive 
"dead end." 

. . .

::: {.poll .nonincremental}
- Did your hypothesis fail, or did you? 

- Did you just waste 3 weeks of work?

- Would you still share your work in a public repo?
:::

```{r}
#| echo: false
countdown::countdown(4, bottom = 0)
```


::: {.notes}
1) Did your hypothesis fail, or did you?
It's important to separate the quality of your code from the behavior of nature.  
Also, an automated "failed" experiment is actually a massive engineering success.

2) Did you just waste 3 weeks of work?
The danger of deleting your dead ends: What happens to the scientific community 
when researchers hide their failures, forcing the next person to waste weeks 
re-testing the exact same wrong idea.

3) Would you still share your work in a public repo?
Submit and store it with the exact same reproducibility standards as a success, 
because documenting "blind alleys" saves future researchers from wasting time 
on the same mistake.
:::


## Scenario 4

Your script runs a 4-hour simulation on a large dataset, 
processing everything live in temporary memory to output a final summary paragraph. 
When your professor spots an anomaly and asks to see the data right after 
the initial filtering step, you realize nothing was saved along the way.

. . .

::: {.poll .nonincremental}
- What are some advantages of a single "all-or-nothing" live run?

- What are some disadvantages of saving intermediate steps?
:::

```{r}
#| echo: false
countdown::countdown(4, bottom = 0)
```

::: {.notes}
Disadvantages of a pipeline that only saves the final answer: impossible 
to troubleshoot, audit, or verify when a bug or anomaly inevitably appears 
in the middle.

Discuss the massive waste of time and computational energy that happens 
when you treat data analysis as a single "all-or-nothing" live run 
instead of caching milestones.

Debate the engineering trade-off between consuming hard drive space 
to store temporary milestone files versus consuming hours of processing time 
to re-run scripts.

Better to transform a 4-hour pipeline into a modular, flexible workflow 
that can be paused, inspected, and restarted at any point.
:::


# Quiz-Poll Time

Open your laptops / phones

## Google Form

\

![](images/poll-01-qr-code.svg){fig-align="center"}

\

[tinyurl.com/2j88yk2n](https://tinyurl.com/2j88yk2n)


## Poll

Be completely honest: When cleaning a dataset or fixing file formatting issues 
for an assignment, how often do you rely on manual copy-pasting, visual cell 
edits, or find-and-replace clicks instead of writing a code script?

A) [100% Code Scripted:]{.bold-hilit} I never touch the raw file manually; 
every transformation is written in a reproducible script lines.

B) [The Hybrid Scramble:]{.bold-hilit} I try to script, but if a 
deadline is tight, I secretly open Excel to quickly delete bad rows or 
copy-paste columns.

C) [Purely Manual:]{.bold-hilit} I do almost all data formatting visually 
using trackpad movements and mouse clicks because it feels faster.

D) [The "I Don't Track It" Stance:]{.bold-hilit} I change cells casually as 
I work and completely lose track of what I manually edited by the end of the day.


## Poll

- You notice a minor discrepancy in plot you generated 3 weeks ago. You need to backtrack 
to the exact code state that produced that specific figure. 
- What is the reality of your current local file system?

A) [Perfect Provenance:]{.bold-hilit} Every single script modification is systematically frozen in a timestamped Git commit hash; I can jump back in seconds.

B) [The "Folder Vault" Cluster]{.bold-hilit} I have copies named `script_v1.R`, `script_v2_fixed.R`, and `script_final_REAL.R`. It will take hours of manual guessing.

C) [The Overwrite Abyss:]{.bold-hilit} I continuously modify a single file without saving historical backups. That past code state is permanently gone.

D) [The Cloud Illusion:]{.bold-hilit} I assume my background Google Drive/Dropbox sync will save me, even though it doesn't track execution states.


## Poll

You run a classmate's random simulation script on your machine 
using identical data inputs. But they forgot to set a static random seed. 
Your output values are *approximately* close (e.g., 84.2% accuracy) but not *exactly* 
equal to their published value (84.5%). What is your scientific verdict?

A) [Acceptable Replication:]{.bold-hilit} Close enough is good enough for 
computational science; minor stochastic fluctuations don't undermine the core theory.

B) [Unverifiable Black Box:]{.bold-hilit} An approximate match is a failure; 
without an exact mathematical match via a random seed, I cannot verify if 
the pipeline is truly bug-free.

C) [Indifferent Stance:]{.bold-hilit} I assume the difference is caused by 
our different operating system kernels or processor architectures anyway, 
so a seed won't fix it.


## Poll

Think about how you write lab reports, group notes, or draft research paragraphs. 
Where do your conceptual interpretations live relative to the raw output data 
files that support them?

A) [Fully Welded Documents:]{.bold-hilit} I use integrated dynamic document 
tools (like Quarto or Jupyter Notebooks) so my text and live data logic are 
locked together natively.

B) [The Absolute Path Blind Bind:]{.bold-hilit} I manually type hardcoded 
server folder paths (e.g., `/Users/name/Desktop/run_3`) inside a separate 
Microsoft Word document.

C) [The Total Disconnect:]{.bold-hilit} My data lives in a folder on a 
local server, and my interpretations live entirely in casual emails, 
Slack chats, or scrap paper notes.


## Poll

You are preparing to submit your finalized computational report. What is your 
group's stance on providing full public access to your custom scripts and 
intermediate datasets?

In exactly **one word**, submit the single biggest psychological or technical 
barrier that makes you hesitate to hit "Public Release" on a dataset.*



# Take Aways

## From today's paper

\

| "As a minimal requirement, you should at least 
| be able to reproduce the results yourself.
| . . .
| Although it may sound like a very
| weak requirement, even this level of
| reproducibility will often require a certain
| level of care in order to be met."
